---
title: "CSC477_Final-project_Oenardi_101924"
author: "Janice Gifano Oenardi"
format: html
css: janice-font.css
---

# Federal Reservce Economic Data Source


**FRED Dataset**  
Our group chose the Federal Reserve Economic Data (FRED) data source, with the link: <https://fred.stlouisfed.org/>, because we would like to delve more into the US economy to conduct economic analysis and financial research. Our group think that the dataset contains lots of information that we can draw meaningful relationships and significance to explore. Additionally, there are a lot of different data sets under different sections, which enable us to explore different aspects of the US economy including: macroeconomic analysis, energy prices and consumption, labor market dynamics, and education spending and outcomes.


**Context of the FRED Dataset for Macroeconomic Analysis**  
The Federal Reserve Economic Data (FRED) is a database maintained by the Research division of the Federal Reserve Bank of St. Louis that has more than 816,000 economic time series from various sources, making it a crucial resource for economists, researchers, and policymakers to analyze economic trends and make informed decisions regarding the U.S. economy; essentially, it's a central repository for a wide range of economic data, allowing users to easily compare and visualize different economic indicators over time.


**Who uses the data?**  
Economists, policy makers, business leaders/owners, researchers, financial market analysts, and students can use the FRED data.

**What does the FRED offer? What does the dataset include?**   
FRED contains frequently updated US macro and regional economic time series at annual, quarterly, monthly, weekly, and daily frequencies. The FRED data source has time series data from mid 1900s and up to the most recent days in 2024. FRED aggregates economic data from a variety of sources, most of which are US government agencies including the Bureau of Labor Statistics, Census Bureau, and other national and international organizations. Some topics that the FRED provide data on include GDP, employment rate, inflation rate, housing price index, trade, and more.  



# Variables 


**Inflation**  

**10-Year Breakeven Inflation Rate**: Represents the market's expectation for the average annual inflation over the next ten years. Break-even inflation is the difference between the nominal yield on a fixed-rate investment and the real yield (fixed spread) on an inflation-linked investment of similar maturity and credit quality. The latest value implies what market participants expect inflation to be in the next 10 years, on average.  
-   **Units:** Percent, Not Seasonally Adjusted  
-   **Frequency:** Daily  
-   **Source:** Federal Reserve Bank of St. Louis  


**National Income**

**Nominal Gross Domestic Product (GPD)**: to track the total market value of all finished goods and services produced within a country's borders in a specific period.  
-   **Source:** International Monetary Fund  
-   **Units:** Millions of Domestic Currency  
-   **Frequency:** Quarterly  
    

**Real Personal Income**: adjusted for inflation, it measures the purchasing power of individuals.  
-   **Source:** U.S. Bureau of Economic Analysis  
-   **Units:** Billions of Chained 2017 Dollars, Seasonally Adjusted Annual Rate  
-   **Frequency:** Monthly  


**International trade**    
-   **Trade Balance: Goods and Services, Balance of Payments Basis**: the difference between a country's value of exports and imports of goods and services over a specific period of time.  
    -   **Source:** U.S. Bureau of Economic Analysis  
    -   **Units:** Millions of Dollars, Seasonally Adjusted  
    -   **Frequency:** Monthly


**Fiscal Policy**  

**Consumer Opinion Surveys: Consumer Prices: Future Tendency for United States**: reflects consumers' level of optimism about the performance of the economy.  
-   **Source:** Organization for Economic Co-operation and Development  
-   **Units:** Percentage balance, Seasonally Adjusted  
-   **Frequency:** Monthly  

**Federal Debt: Total Public Debt as Percent of Gross Domestic Product**: tracks the government's debt level in relation to the GDP (economic output).  
-   **Source:** U.S. Office of Management and Budget  
-   **Units:** Percent of GDP, Seasonally Adjusted  
-   **Frequency:** Quarterly  


**Federal Government Current Receipts**: tracks the total money the federal government obtains from taxes.  
-   **Source:** U.S. Bureau of Economic Analysis  
-   **Units:** Billions of Dollars, Seasonally Adjusted Annual Rate  
-   **Frequency:** Quarterly  


**Employment levels**  
**Unemployment Rate**: the unemployment rate represents the number of unemployed as a percentage of the labor force. Labor force data are restricted to people 16 years of age and older, who currently reside in 1 of the 50 states or the District of Columbia, who do not reside in institutions (e.g., penal and mental facilities, homes for the aged), and who are not on active duty in the Armed Forces.  
-   **Source:** U.S. Bureau of Labor Statistics    
-   **Units:** Percent, Seasonally Adjusted  
-   **Frequency:** Monthly  



# Significance of the dataset


**Where is the data from?**  
The data is taken from reputable organizations such as the U.S. Bureau of Labor Statistics (BLS), Bureau of Economic Analysis (BEA), Federal Reserve Board, World Bank, and other national and global reputable organizations.  


**Why is the FRED important?**  

The FRED offers reliable, intensive, and broad database for analyzing and visualizing key economic indicators, which can help make data-driven and more informed decisions regarding business strategies, investment, and monetary/public policy.  



**Significance in the real world**

Inflation affects purchasing power of consumers, interest rates, wages, and more. Analyzing inflation can help to anticipate inflationary pressures, which would affect the whole economy in various industries. Secondly, the national income measures the overall health of a country financially by providing insight onto the citizen's economic well-being. Additionally, the fiscal policy variables influence stability and growth by affecting tax income, debt level, government spending, investments, and more. Furthermore, employment levels are a key indicator of the economic health, as it influences consumer spending and business investments. Lastly, international trade shows the country's global economic imports and exports and helps to understand a country's competitiveness.

# Loading the dataset

```{r}
##install.packages("readxl")  # Install if not already installed
library(readxl)
```

```{r}
# Replace 'your_file.xlsx' with the path to your Excel file
mydata <- read_excel("Final_Project_copy_2.xls")

# View the first few rows of the data
head(mydata)
str(mydata)
```

# GDP Growth and Inflation Expectations

Exploring the relationship between GDP growth and inflation expectations can provide insights into the overall economic stability and help policymakers predict future economic conditions, guiding decisions on interest rates, fiscal policies, and inflation control measures.


**Steady Growth in U.S. GDP Trend over the Past 20 Years**  

```{r}
## Load the required packages
# install.packages("ggplot2")
library(ggplot2)
# install.packages("dplyr")
library(dplyr)
library(scales)

## Renaming variables for simplicity
mydata <- mydata %>%
  rename(Year = `Years (DATE)`,
         Inflation_Rate = `Average of 10-Year Breakeven Inflation Rate`,
         GDP = `Average of Nominal Gross Domestic Product for United States`,
         Consumer_Prices = `Average of Consumer Opinion Surveys: Consumer Prices: Future Tendency for United States`,
         Unemployment_Rate = `Average of Unemployment Rate`,
         Trade_Balance = `Average of Trade Balance: Goods and Services, Balance of Payments Basis`,
         Personal_Income = `Average of Real Disposable Personal Income`,
         Government_Receipts = `Average of Federal Government Current Receipts`,
         Federal_Debt_Ratio = `Average of Federal Debt: Total Public Debt as Percent of Gross Domestic Product`)


## Converting 'Year' column from character to numeric
mydata$Year <- as.numeric(mydata$Year)

## Preview the dataset
head(mydata)
str(mydata)

# Create the line graph
ggplot(data = mydata, aes(x = Year, y = GDP)) +
  geom_line(color = "lightblue", size = 1) +
  geom_point(color = "blue", size = 1) +  # Add points on the line
  labs(title = "GDP Trend in the US (Last 20 Years)",
       x = "Year",
       y = "Nominal GDP (in USD)") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()

# Define economic downturn periods (example years)
recessions <- data.frame(
  start = c(2008, 2019),  # Start of recession
  end = c(2009, 2020)     # End of recession
)

# Plot the graph with shaded areas for recessions
ggplot(data = mydata, aes(x = Year, y = GDP)) +
  geom_line(color = "lightblue", size = 1) +
  geom_point(color = "lightblue", size = 2) +  # Add points on the line
  geom_rect(data = recessions, aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf),
            fill = "grey", alpha = 0.2, inherit.aes = FALSE) +  # Shaded areas
  labs(title = "GDP Trend in the US (Last 20 Years)",
       x = "Year",
       y = "Nominal GDP (in USD)") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()
```

The graph shows the trend of GDP in the U.S. over the last 20 years. The blue line represents the average annual changes in GDP, while the shaded grey areas highlight two significant periods in U.S. history, which are the 2008 financial crisis and the 2019 COVID-19 pandemic. Overall, the graph visualize how resilient the U.S. economy is despite the face of these challenges, as the GDP appears to continue increasing over the years. In the past 4 years, the GDP has been rising more signfiicantly at a higher pace, as seen from the stepper upward line trend. The significant increase in average GDP year over year indicate \_\_

### General trend:

The overall trend of the U.S. GDP shows an upward trend, showing that the economy has been growing significantly throughout the past two decades. The graph shows the periods of expansion and minor dips in GDP, but the overall growth appears steady.

### Economic downturns:

There are two shaded areas around 2008 and 2019. The first shaded area around 2008 makrs hte global financial crisis. During this period, the GDP dipped, but by 2010, it started to recover. The other shaded area around 2019 shows the impact of the COVID-19 pandemic. There was a downturn in GDP, however quickly bounced back in the following years.

### Overall conclusion

Overall, the graph visualizes how resilient the U.S. economy is despite the face of these challenges, as the GDP appears to continue increasing over the years. In the past four years, the GDP has been rising more significantly at a higher pace, as seen from the steeper upward line trend. The significant increase in average GDP year over year indicates a robust economic recovery, that is driven by strong consumer spending, increase in investment, and favorable fiscal policies that support growth.

# Does Higher Inflation Mean Higher GDP?

```{r}
ggplot(mydata, aes(x = GDP, y = Inflation_Rate, color = GDP)) +
  geom_point(aes(size = Inflation_Rate), alpha = 0.7) +
  labs(title = "Scatter Plot of GDP vs Inflation Rate",
       x = "GDP (in dollars)",
       y = "Inflation Rate (%)") +
  scale_x_continuous(labels = comma) +  # Format x-axis with commas
  scale_color_gradient(low = "lightblue", high = "darkblue", name = "GDP Value") +
  theme_minimal()
```

The scatter plots shows the relationship between GDP (in USD) and inflation rate (%). The color of the points represent the GDP value (with lower values being bright green and higher values being light blue), while the size of the poitns reflect the inflation rate (with smaller points showing a lower rate and larger points showing a higher rate).

### Unclear correlation between GDP and Inflation Rate

The scatterplot does not seem to show a strong clear pattern between GDP and inflation rate. The points spread across different GDP values, without a clear upward or downward trend. This could suggest that GDP and inflation are not strongly correlated as they don't have a direct, simple relationship.

### High GDP values do not always align with high Inflation Rates, and vice versa

However, it is noticable that the points with higher inflation rates are scattered across both high and low GDP values, which may indicate that even as GDP changes, the inflation does not necessarily flow a predictable path. In other words, high GDP does not directly align with high inflation rates, and vice versa.

### Overall conclusion

In summary, this graph shows that there isn’t a straightforward relationship between GDP and Inflation Rate. The factors affecting inflation may be more complex and not directly tied to GDP alone.

## Steady Increase in Disposable Income vs. Volatility in Consumer Confidence in Prices 

```{r}
scale_factor <- max(mydata$Personal_Income) / max(mydata$Consumer_Prices)

ggplot(mydata, aes(x = Year)) +
  geom_line(aes(y = Personal_Income, color = "Personal_Income"), size = 1) +
  geom_point(aes(y = Personal_Income, color = "Personal_Income"), size = 2) +
  geom_line(aes(y = Consumer_Prices * (max(Personal_Income) / max(Consumer_Prices)),
                color = "Consumer_Prices"), size = 1) +
  geom_point(aes(y = Consumer_Prices * (max(Personal_Income) / max(Consumer_Prices)),
                color = "Consumer_Prices"), size = 2) +
  scale_y_continuous(
    name = "Disposable Personal Income (in USD)",
    sec.axis = sec_axis(~ . / scale_factor, name = "Consumer Confidence in Prices (%)")
  ) +
  scale_color_manual(values = c("Personal_Income" = "lightblue", "Consumer_Prices" = "lightgreen")) +
  labs(title = "Disposable Personal Income Vs. Consumer Confidence in Prices",
       x = "Year") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(plot.title = element_text(hjust = 0.5))
```

### Steady Increase in Disposable Personal Income
The disposable personal income shows a generally stead upward trend, which indicates that people's purchasing power in the U.S. generally increased in the last 20 years. This could be attributed to economic growth, inflation, wage increases, and other adjustments over the years. However, there is a significant dip in the disposable personal income after 2020, which is most likely attributable to the economic shock caused by the COVID-19 pandemic that disrupted earnings and employment for many households. The subsequent quick recovery in the following years signify the government's efficient policies and support programs that are able to boost the people's personal incomes.

### Volatility in Consumer Confidence in Prices
The green line, which represents the percentage of consumer optimism regarding prices, reflects hoow the prices have changed over the past 20 years, likely due to inflation, supply-demand imbalances, and other economic disruptions. There's a significant spike after 2020, during the same time time disposable personal income decreased. This could correspond to the economic effects of the COVID-19 pandemic that caused shortages, inflation, and widespread disruptions to supply chains and consumer markets globally.

## High Consumer Confidence in Prices in the Past Four Years

```{r}
ggplot(mydata, aes(x = factor(Year), y = 1, fill = Consumer_Prices)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Consumer Confidence in Prices") +
  labs(title = "Consumer Confidence in Prices Heatmap Over Time",
       x = "Year",
       y = "") +
  theme_minimal() +
  # To organize the x-axis so everything looks neat
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5))
```

### Consecutively Low Consumer Confidence is Followed by High Consumer Confidence in 2021-2024
The heatmap shows varying shades of blue, with darker shades indicating higher levels of Consumer Confidence in Prices. There are lighter colros in earlier years, which may suggest that consumer confidence level in those periods were relatively low. On the other hand, there are some darker blue bands towards the more recent years, indicating periods of higher consumer confidence.

### Notable Dip Around 2008 and Signficant Spike After 2020 in Consumer Confidence in Prices
There is a noticable dip from 2008 to 2009, with the color changing from darker blue to lighter blue, which may correlate with the global financial crisis affecting consumer confidence in prices. This may be due to economic instability, which caused consumers to have low confidence levels in prices, especially due to the overall decline in economic activity after the housing market crisis.  After 2020, there is a sharp increase in intensity, suggesting a spike in Consumer Confidence in Prices, which might be due to economic responses to the COVID-19 pandemic and recovery measures. The dark blue for the years 2022-2024 may imply that consumer confidence in prices has been quite strong recently and quite consistently, potentially due to post-pandemic recovery measures, which include economic policies, stimulus packages, or other factors affecting consumer confidence. In overall, the color intensity changes may reflect broader economic cycles, where confidence generally decreases during economic downturns (e.g., 2008) and increases during recovery periods (post-2020). 


## Upward Trend in Government Receipts and Federal Debt Ratio

```{r}
scale_factor <- max(mydata$Government_Receipts) / max(mydata$Federal_Debt_Ratio)

# Create dual chart
ggplot(mydata, aes(x = Year)) +
  # Bar chart for Government Receipts
    geom_bar(aes(y = Government_Receipts), 
           stat = "identity", 
           fill = "#48c9b0",      # Color for the bars
           width = 0.7) +         # Adjusting bar width
  # Line chart for Federal Debt Ratio with secondary axis
  geom_line(aes(y = Federal_Debt_Ratio * scale_factor, color = "Federal Debt Ratio"), size = 1) +
  geom_point(aes(y = Federal_Debt_Ratio * scale_factor, color = "Federal Debt Ratio"), size = 2)  +
  scale_y_continuous(
    name = "Federal Government Receipts (in billions $)",
    sec.axis = sec_axis(~ . / scale_factor, name = "Federal Debt as a Percent of GDP (%)")
  ) +
  labs(title = "Government Receipts (in billions $) and Debt as a Percent of GDP (%)",
       x = "Year") +
  scale_color_manual(name = "", values = c("Federal Debt Ratio" = "#1f78b4")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels for readability
  theme(plot.title = element_text(hjust = 0.5))
```
### Upward Trend in Government Receipts

There is a general upward trend in government receipts over the observed period, indicating a steady increase in the amount of revenue collected by the federal government. Notable dips in receipts can be observed around 2008-2010, which coincides with the Great Recession, a period of economic downturn when tax revenues typically decline. After this period, government receipts show a recovery and continue to rise, reaching a peak around 2022-2024.

### Uppward trend with More Fluctuations in Federal Debt Ratio

The Federal Debt Ratio also shows an upward trend, but with more fluctuations compared to receipts.
The most noticeable increase in the debt ratio occurs around the 2008 financial crisis, where the federal government likely took on more debt as part of economic stimulus efforts.
Another sharp increase is visible around 2020, likely due to the COVID-19 pandemic, when government debt rose significantly to support the economy through various relief programs.
After 2021, the debt ratio seems to stabilize but remains high compared to earlier years.


## Government Receipts is not Directly Correlated to Debt Ratio

Although both federal receipts and debt as a percentage of GDP generally increase over time, they are not directly synchronized. During periods of economic crises (2008 and 2020), government receipts appear to to dip or stabilize, while debt as a percent of GDP rises sharply.
After these periods, government receipts gradually increase, but the debt ratio remains elevated, indicating that higher revenue does not immediately reduce the debt-to-GDP ratio.

The long-term trend suggests that even though government revenue increased, it has not increased within the same pace with the increase in debt relative to the GDP. This hihlghts potential concerns about the sustainability of the debt, as it appears to increase more significantly than  the total money the federal government obtains from taxes.  



## Fluctuating Unemployment Rate in US

```{r}
# Create the line graph
ggplot(data = mydata, aes(x = Year, y = Unemployment_Rate)) +
  geom_line(color = "lightblue", size = 1) +
  geom_point(color = "blue", size = 1) +  # Add points on the line
  labs(title = "Unemployment Rate in the US (Last 20 Years)",
       x = "Year",
       y = "Unemployment Rate (in %)") +
  theme_minimal()


# Define economic downturn periods (example years)
recessions <- data.frame(
  start = c(2007, 2019),  # Start of recession
  end = c(2009, 2020)     # End of recession
)

# Plot the graph with shaded areas for recessions
ggplot(data = mydata, aes(x = Year, y = Unemployment_Rate)) +
  geom_line(color = "lightblue", size = 1) +
  geom_point(color = "lightblue", size = 2) +  # Add points on the line
  geom_rect(data = recessions, aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf),
            fill = "grey", alpha = 0.2, inherit.aes = FALSE) +  # Shaded areas
  labs(title = "Unemployment Rate in the US (Last 20 Years)",
       x = "Year",
       y = "Unemployment Rate (in %)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

### Unemployment Rate Appears to Surge in Periods of Recession
As shown in the graph, there are two periods in time where unemployment rate spikes significantly, which started in 2007 and peaked in 2010. The second period of time where unemployemnt rate spiked starts in 2019 and peaked in 2020. It is evident that the 2008 economic crisis and COVID-19 pandemic caused the unemployment rate to increase. During high uenmployment levels, people have less money to spend, so demand decreases, and prices (inflation) stay low. Wages also tend to remain stagnant. However, the spike in unemployment rate is followed by a continous and steady decrease in unemployment rate, showing that the economy is recovering. This indicates that as unemployment rate decreases, inflation start to increase for periods after 2010.

## 2008 Economic Crisis Has Bigger Effect on Unemployment Rate than COVID-19 
The graph shows that the recovery after the 2007-2008 economic crisis in the US is longer than the recover after the COVID-19 pandemic.The 2008 financial crisis was an epic financial and economic collapse that cost many ordinary people their jobs, their life savings, their homes, or all three. Therefore, the effects of the crisis caused unemployment rate to spike significantly (reaching over 9%), more than COVID-19 caused unemployment rate to increase (slightly above 8%). The amount of years it took for the unemployment rate to decrease are 9 years, from 2011 to 2019. The decreasing levels of unemployment rates signify that inflation starts to increase during those years. More jobs and higher wages increase household incomes and lead to a rise in consumer spending, further increasing aggregate demand and the scope for firms to increase the prices of their goods and services. 

## Effect of Unemployment Rate to GDP

```{r}
scale_factor <- max(mydata$GDP) / max(mydata$Unemployment_Rate)

# Create dual chart
ggplot(mydata, aes(x = Year)) +
  # Bar chart for GDP
    geom_bar(aes(y = GDP), 
           stat = "identity", 
           fill = "#48c9b0",      # Color for the bars
           width = 0.7) +         # Adjusting bar width
  # Line chart for Unemployment RAte with secondary axis
  geom_line(aes(y = Unemployment_Rate * scale_factor, color = "Unemployment_Rate"), size = 1) +
  geom_point(aes(y = Unemployment_Rate * scale_factor, color = "Unemployment_Rate"), size = 2) +
  scale_y_continuous(
    name = "GDP (in millions $)",
    sec.axis = sec_axis(~ . / scale_factor, name = "Unemployment Rate (%)")
  ) +
  labs(title = "GDP (in millions $) and Unemployment Rate (%)",
       x = "Year") +
  scale_color_manual(name = "", values = c("Unemployment_Rate" = "#1f78b4")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate x-axis labels for readability
  theme(plot.title = element_text(hjust = 0.5))
```

### Spikes in Unemployment Rates Leads to Decrease in GDP
This chart illustrates the relationship between GDP (in millions of dollars) and the unemployment rate (%) over time, highlighting how spikes in unemployment rates often correspond with subsequent declines or slower growth in GDP. For example, the significant rise in the unemployment rate around 2009 aligns with the economic downturn during the financial crisis, which visibly impacted GDP growth. Similarly, the spike in unemployment in 2020 due to the COVID-19 pandemic coincides with a decline in GDP growth, indicating the strong influence that high unemployment rates have on economic productivity and output. This pattern shows the negative impact of rising unemployment on economic health, as reduced consumer spending and decreased business activity contribute to slower GDP growth during these periods.

### Economic Resilience Amidst Fluctuating Unemployment Rates
The graph suggests that despite spikes in unemployment, U.S. GDP demonstrates a trend of steady growth over the long term. Even after significant increases in unemployment and economic downturns, GDP eventually resumes its upward trajectory, highlighting the economy’s resilience and capacity for recovery. The pattern shows that while unemployment has short-term impacts on economic output, other factors, such as policy responses, innovation, and investment, play crucial roles in sustaining and driving GDP growth over time. This resilience points to the adaptability of the economy, as it manages to recover and expand even after periods of high unemployment.

## What is the Trend in U.S. Trade Balance?

```{r}
# Create the line graph
ggplot(data = mydata, aes(x = Year, y = Trade_Balance)) +
  geom_line(color = "lightblue", size = 1) +
  geom_point(color = "lightblue", size = 2) +  # Add points on the line
  labs(title = "U.S. Trade Balance in millions of $ (Last 20 Years)",
       x = "Year",
       y = "Trade Balance (in millions of $)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

### Several Fluctuations in U.S. Trade Balance 
This graph illustrates the U.S. trade balance in millions of dollars over the last 20 years. The trade balance, which represents the difference between exports and imports, has consistently remained in a deficit, indicating that the U.S. imported more goods and services than it exported during this period. Between 2005 and 2010, the trade deficit fluctuated significantly, reaching its peak around 2009, likely due to the economic impact of the Global Financial Crisis. This period saw a temporary improvement in the deficit, potentially due to reduced imports driven by decreased consumer spending during the recession. However, the deficit resumed its downward trend in subsequent years as economic recovery led to increased demand for imports.

### Deepening of U.S. Trade Deficit
In the last decade, the trade deficit has deepened considerably, with notable spikes in 2020 and 2021. The sharp decline in the trade balance during 2020 may reflect disruptions caused by the COVID-19 pandemic, such as shifts in global supply chains and surging imports of medical and consumer goods. The temporary improvement in 2021 suggests an adjustment as the economy began to recover. Despite these fluctuations, the long-term trend points to an increasing reliance on imports relative to exports, highlighting structural challenges in U.S. trade policy and the global competitiveness of domestic industries. Addressing this deficit will likely require targeted strategies to boost exports and reduce dependency on imported goods.




## Bubble Chart Analysis 
```{r}
ggplot(mydata, aes(
  x = Trade_Balance,
  y = GDP,
  size = Federal_Debt_Ratio
)) +
  geom_point(alpha = 0.7, color = "steelblue") +
  scale_y_continuous(labels = scales::comma) +
  scale_size_continuous(name = "Federal Debt Level", range = c(2, 10)) +
  labs(
    title = "Trade Activity vs. Economic Indicators",
    x = "Trade Balance (in billion $)",
    y = "GDP (in millions $)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
  )

```

### Trade Deficit and Economic Growth Correlation
The graph reveals an intriguing relationship between trade balance and GDP, suggesting that countries with larger trade deficits (more negative trade balance) tend to exhibit higher GDP values. This counterintuitive observation indicates that trade deficits may not necessarily hinder economic growth but could be reflective of high domestic demand for imports, which often accompanies robust economic activity. Notably, GDP levels cluster between $4 trillion and $7 trillion, even as the trade balance ranges significantly from -$80 billion to -$40 billion.

### Federal Debt as an Amplifier of Economic Dynamics
The size of the bubbles, representing federal debt levels, highlights another layer of complexity. Larger debt levels (bubbles) tend to align with countries that have higher GDP and significant trade deficits, suggesting that borrowing may play a pivotal role in sustaining both high GDP and large trade imbalances. This pattern underscores the intricate interplay between national borrowing, trade policies, and economic performance. It raises questions about the sustainability of this model and whether such nations rely on federal debt to fuel economic activity and finance their trade gaps.

