---
title: "Week 3 Deliverable"
author: "Diego J Mendoza Future Ph.D"
format: html
---

# Week 3 deliverables

## Foucs

Week 3: Industrial Energy Consumption and Price Sensitivity
Deliverables:
Visualization 1: A scatter plot comparing natural gas prices and industrial energy consumption to assess how sensitive industries are to price changes.
Visualization 2: A bar chart showing energy consumption by industry, identifying which sectors are most energy-dependent.


### Visualization 1 Scatterplot

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(ggpubr)

```

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(ggpubr)

# Load datasets
natural_gas_data <- read.csv("PNRGINDEXM.csv", header = TRUE)
gas_pricing <- read.csv("NATURALGASD11.csv", header = TRUE)

# Ensure "DATE" is in date format
natural_gas_data$DATE <- as.Date(natural_gas_data$DATE, format = "%Y-%m-%d")
gas_pricing$DATE <- as.Date(gas_pricing$DATE, format = "%Y-%m-%d")

# Merge datasets on DATE column
merged_data <- natural_gas_data %>%
  inner_join(gas_pricing, by = "DATE")

```

```{r}
# Ensure the columns are numeric and remove any rows with NA values
merged_data$NATURALGASD11 <- as.numeric(merged_data$NATURALGASD11)
merged_data$PNRGINDEXM <- as.numeric(merged_data$PNRGINDEXM)

# Remove rows with NA values
merged_data <- na.omit(merged_data)

```


```{r}
# Scatter plot with Natural Gas Prices and Global Energy Index
ggscatter(merged_data, x = "NATURALGASD11", y = "PNRGINDEXM", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "Natural Gas Prices", ylab = "Global Energy Index") +
  labs(title = "Scatterplot of Natural Gas Prices vs. Global Energy Index")


```

```{r}
# Scatter plot with every other x-axis label removed
ggscatter(merged_data, x = "NATURALGASD11", y = "PNRGINDEXM", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "Natural Gas Prices", ylab = "Global Energy Index") +
  labs(title = "Scatterplot of Natural Gas Prices vs. Global Energy Index") +
  scale_x_continuous(breaks = seq(min(merged_data$NATURALGASD11), 
                                  max(merged_data$NATURALGASD11), by = 30)) +
  scale_y_continuous(breaks = seq(min(merged_data$PNRGINDEXM),
                                  max(merged_data$PNRGINDEXM), by = 15)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
#### Staticics

R score = 0.34,
P value = 0.1,

P and R value indicate a weak positive correlation between the variables.
The P value indicate that any correlation could be because of chance rather than causation.

From this scatterplot of Natural Gas Prices vs. Global Energy Index, with an 
𝑅
R score of 0.34 and 
𝑝
p-value of 0.1, we can make the following observations:

Correlation (R = 0.34):

The 
𝑅
R score of 0.34 represents a weak positive correlation between natural gas prices and the global energy index. This suggests that as natural gas prices increase, the global energy index tends to increase as well, but only slightly.
Since 0.34 is relatively low, the relationship between these two variables is not strong.
Statistical Significance (p = 0.1):

The 
𝑝
p-value of 0.1 indicates that this correlation is not statistically significant at the typical significance levels (usually 
𝑝
<
0.05
p<0.05).
This suggests that the observed correlation might be due to random chance rather than a meaningful association between natural gas prices and the global energy index.
Confidence Interval:

The gray shaded area represents the confidence interval for the regression line, which appears relatively wide. This further indicates a high degree of uncertainty in predicting the global energy index from natural gas prices, given the weak correlation.
Summary
The scatterplot shows a weak positive relationship between natural gas prices and the global energy index, but the association is not statistically significant. This suggests that changes in natural gas prices may not be a reliable predictor of changes in the global energy index.








###Visualizationn
```{r}

sector_energy_spending <- read.csv("CPGREN01USM659N.csv")



```

```{r}
# Load required libraries
library(ggpubr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)

# Load data
sector_energy_spending <- read.csv("CPGREN01USM659N.csv")

# Convert DATE to Date format and CPGREN01USM659N to numeric
sector_energy_spending$DATE <- as.Date(sector_energy_spending$DATE)
sector_energy_spending$CPGREN01USM659N <- as.numeric(sector_energy_spending$CPGREN01USM659N)

# Extract the year and create a decade column
sector_energy_spending <- sector_energy_spending %>%
  mutate(Year = as.numeric(format(DATE, "%Y")),
         Decade = paste0(floor(Year / 10) * 10, "s"))

# Create the boxplot grouped by Decade
ggboxplot(sector_energy_spending, x = "Decade", y = "CPGREN01USM659N",
          color = "Decade", palette = "jco",
          ylab = "Sector Energy Spending", xlab = "Decade",
          title = "Boxplot of Sector Energy Spending by Decade") +
  theme_minimal() +
  theme(legend.position = "none")


```
#### Statics

General Trend:

The median sector energy spending appears to fluctuate across decades, with notable increases in the 1970s, 2000s, and 2020s.
The distribution of spending varies significantly, with wider interquartile ranges (IQRs) in the 1970s, 2000s, and 2020s, indicating higher variability in these decades.
Outliers:

There are multiple outliers, especially in the 1970s, 1980s, and 2000s, suggesting instances where energy spending was unusually high or low relative to the general spending range in each of these decades.
For example, a high outlier in the 1970s is above 30, and low outliers in the 2000s drop below -10.
Median Values:

The median energy spending generally hovers around or above zero in most decades.
In the 2020s, the median spending is notably higher than previous decades, which could indicate an increase in energy spending in recent years.
Interquartile Range (IQR):

The IQR is relatively narrow in the 1950s and 1960s, suggesting low variability in spending during those decades.
However, the IQR widens in the 2000s and 2020s, indicating that spending became more volatile in these periods.
Skewness:

Some decades, like the 2000s and 2010s, show an asymmetrical distribution, with longer tails on one side, suggesting skewness in the data.
Summary
Sector energy spending varies significantly across decades, with notable volatility in recent years (2000s and 2020s).
Outliers in the 1970s, 1980s, and 2000s highlight unusual spending spikes or drops.
The upward shift in the median in the 2020s suggests a potential increase in energy spending in recent times.
