---
title: "Week2Deliverables"
author: "Diego J Mendoza"
---

# Week 2 Deliverables

### Deliverables

Visualization 1: A stacked area chart comparing household energy consumption by type (e.g., electricity, natural gas) over time.

Visualization 2: A line graph of the Global Price of Energy Index against electricity prices. This explores how overall energy prices influence electricity costs.

### Background Information

EnergyIndexUrbanHouseholdCost (CUSR0000SAH21): This index tracks monthly changes in the cost of household energy (e.g., electricity, natural gas) for urban consumers across the United States. It's seasonally adjusted, providing a clear view of inflation trends in residential energy. 

ConsumerPriceEnergyUSA (CPGREN01USM659N): This dataset shows the consumer price index for energy, highlighting inflation in energy costs specific to the U.S. It measures yearly changes, offering insights into the overall price trend for energy. 

NaturalGasConsumptionMonthlySeries (NATURALGASD11): This dataset captures monthly natural gas consumption, providing information useful for tracking seasonal and annual trends in household and industrial energy use. 

ResidentialEnergyInflationTrend (DELCRC1A027NBEA): This series reports inflation trends within residential energy, tracking the economic impact of rising energy costs on household expenses.


#### Vizulization 1

```{r}
# Load required libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales) # for formatting

# Load the datasets (adjust file paths as needed)
electricity_data <- read.csv("PCU221110221110P.csv")
natural_gas_data <- read.csv("PNRGINDEXM.csv")

# Rename columns for clarity
colnames(electricity_data) <- c("DATE", "Electricity_Consumption")
colnames(natural_gas_data) <- c("DATE", "NaturalGas_Consumption")

# Convert DATE column to Date type
electricity_data$DATE <- as.Date(electricity_data$DATE)
natural_gas_data$DATE <- as.Date(natural_gas_data$DATE)

# Ensure consumption columns are numeric
electricity_data$Electricity_Consumption <- as.numeric(electricity_data$Electricity_Consumption)
natural_gas_data$NaturalGas_Consumption <- as.numeric(natural_gas_data$NaturalGas_Consumption)

# Merge datasets by DATE
merged_data <- merge(electricity_data, natural_gas_data, by = "DATE")

# Reshape data to long format for ggplot stacking
stacked_data <- merged_data %>%
  pivot_longer(cols = c("Electricity_Consumption", "NaturalGas_Consumption"),
               names_to = "Energy_Type", values_to = "Consumption")

# Create the stacked area chart
ggplot(data = stacked_data, aes(x = DATE, y = Consumption, fill = Energy_Type)) +
  geom_area(alpha = 0.8, linewidth = 0.5, color = "white") +  # White borders between layers
  scale_fill_manual(values = c("#FDB462", "#80B1D3")) +  # Custom color palette
  labs(title = "Household Energy Consumption by Type Over Time",
       x = "Date", y = "Consumption",
       fill = "Energy Type") +
  theme_minimal(base_size = 15) +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 20, face = "bold"),
    legend.position = "bottom"
  ) +
  scale_y_continuous(labels = comma)  # Format y-axis with commas

```

#####Insights

**Professional Analysis of Household Energy Consumption Trends**

The stacked area chart illustrates the trends in household energy consumption for electricity and natural gas from 2004 to 2023. Observing the data, three distinct peaks in electricity consumption emerge around 2008, 2014, and especially 2022. These fluctuations align with broader socio-economic, technological, and environmental trends, offering valuable insights into shifts in household energy use over time. Understanding these patterns is essential for professionals in data visualization, as it highlights the potential drivers behind observed data points and the importance of contextual awareness in data interpretation.

The peak in electricity consumption in 2008 could correspond to behavioral shifts during the global financial crisis, when many households reduced discretionary spending and spent more time at home, possibly leading to increased residential energy usage. The following years saw the adoption of energy-efficient appliances and smart technologies, which could explain the decline in consumption as households progressively became more energy-conscious. Such improvements in energy efficiency align with a growing awareness of sustainability practices, demonstrating how technological advancements can directly impact energy demand.

The significant spike in 2022 likely reflects the impact of the COVID-19 pandemic and related changes in daily life. Remote work and schooling increased residential energy demand as more people relied on home heating, cooling, and digital devices. Additionally, the growing adoption of electric vehicles and home offices contributed to rising electricity use in residential areas. Environmental factors, particularly extreme weather patterns driven by climate change, also likely played a role in amplifying demand for heating and cooling systems, especially during periods of extreme temperatures in 2022. These insights underscore the importance of considering economic, technological, and environmental factors when interpreting energy consumption data, as each factor can independently and collectively drive significant changes in household energy use.


#### Vizulization 2

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales) # for formatting

df_energy_index <- read.csv("EnergyIndexUrbanHouseholdCost.csv")
df_cpi_energy <- read.csv("ConsumerPriceEnergyUSA.csv")
df_natgas_consumption <- read.csv("ResidentialEnergyInflationTrend.csv")
df_res_energy_inflation <- read.csv("ResidentialEnergyInflationTrend.csv")

merged_data <- merge(df_energy_index, df_cpi_energy, by = "DATE", all = TRUE)
merged_data <- merge(merged_data, df_natgas_consumption, by = "DATE", all = TRUE)
merged_data <- merge(merged_data, df_res_energy_inflation, by = "DATE", all = TRUE)

merged_data$DATE <- as.Date(paste0(merged_data$DATE, "-01-01"), format = "%Y-%m-%d")

# Drop rows with NA in DATE (optional, but recommended for plotting)
merged_data <- merged_data %>% drop_na(DATE)

colnames(merged_data) <- c("DATE", "Energy_Index", "CPI_Energy", "NatGas_Consumption", "Res_Energy_Inflation")

# Plotting the data
ggplot(data = merged_data, aes(x = DATE)) +
  geom_line(aes(y = Energy_Index, color = "Energy Index")) +
  geom_line(aes(y = CPI_Energy, color = "CPI Energy")) +
  geom_line(aes(y = NatGas_Consumption, color = "Natural Gas Consumption")) +
  geom_line(aes(y = Res_Energy_Inflation, color = "Residential Energy Inflation")) +
  scale_color_manual(values = c("Energy Index" = "blue", "CPI Energy" = "red",
                                "Natural Gas Consumption" = "green", "Residential Energy Inflation" = "purple")) +
  labs(title = "Trends in Energy Index and Consumption over Time",
       x = "Date",
       y = "Index / Consumption / Inflation Rate",
       color = "Legend") +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  theme_minimal()

# Rename columns for merging


#merged_data <- reduce(list(df_energy_index, df_cpi_energy, df_natgas_consumption, df_res_energy_inflation), 
                    #  full_join, by = "DATE")

#merged_data <- merge(df_energy_index, df_cpi_energy, df_natgas_consumption, df_res_energy_inflation , by = "DATE")
```


```{r}
library(ggpubr)

# Read data
data <- read.csv("NaturalGasConsumptionMonthlySeries.csv")

# Convert DATE to Date format and ensure NATURALGASD11 is numeric
data$DATE <- as.Date(data$DATE, format="%m/%d/%Y")
data$NATURALGASD11 <- as.numeric(data$NATURALGASD11)

# Check for any NA values that could arise from non-numeric values
if(any(is.na(data$NATURALGASD11))) {
  warning("There are NA values in the Consumption column. These will be ignored in the plot.")
}

# Plot with ggline
ggline(data, x = "DATE", y = "NATURALGASD11",
       add = "mean_se",  # Add mean line with error bars
       color = "blue", size = 1.2, 
       title = "Natural Gas Consumption Over Time",
       xlab = "Date", ylab = "Consumption") +
  theme_minimal() +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") # Yearly date ticks


```


```{r}
library(ggpubr)

# Read data
data <- read.csv("NaturalGasConsumptionMonthlySeries.csv")

# Convert DATE to Date format
data$DATE <- as.Date(data$DATE, format="%m/%d/%Y")

# Plot line with ggline
ggline(data, x = "DATE", y = "NATURALGASD11",
       add = "mean_se",  # Add mean line with error bars
       color = "blue", size = 1.2, 
       title = "Natural Gas Consumption Over Time",
       xlab = "Date", ylab = "Consumption") +
  theme_minimal() +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") # Yearly date ticks


```

```{r}
library(ggpubr)

# Plot scatter with smooth line
ggscatter(data, x = "DATE", y = "NATURALGASD11",
          add = "reg.line",  # Add regression line
          conf.int = TRUE,   # Add confidence interval
          color = "blue", size = 1.2,
          title = "Natural Gas Consumption with Trend Over Time",
          xlab = "Date", ylab = "Consumption") +
  theme_minimal() +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") # Yearly date ticks


```
```{r}
library(ggpubr)
library(dplyr)

# Read and preprocess data
data <- read.csv("NaturalGasConsumptionMonthlySeries.csv")

# Convert DATE to Date format and ensure NATURALGASD11 is numeric
data$DATE <- as.Date(data$DATE, format="%m/%d/%Y")
data <- data %>%
  mutate(NATURALGASD11 = as.numeric(NATURALGASD11)) %>%
  filter(!is.na(NATURALGASD11))  # Remove rows with NA in NATURALGASD11

# Plot scatter with smooth line, formatting the x-axis to show only the last two digits of the year
ggscatter(data, x = "DATE", y = "NATURALGASD11",
          add = "reg.line",      # Add regression line
          conf.int = TRUE,       # Add confidence interval
          color = "blue", size = 1.2,
          title = "Natural Gas Consumption with Trend Over Time",
          xlab = "Date", ylab = "Consumption") +
  theme_minimal() +
  scale_x_date(date_labels = "%y", date_breaks = "1 year") # Show only last two digits of the year


```
```{r}
library(ggpubr)
library(dplyr)

# Aggregate data by year to get mean consumption per year
data <- read.csv("NaturalGasConsumptionMonthlySeries.csv")
data$DATE <- as.Date(data$DATE, format="%m/%d/%Y")
data$YEAR <- format(data$DATE, "%Y")  # Extract year as a separate column
data <- data %>%
  mutate(NATURALGASD11 = as.numeric(NATURALGASD11)) %>%
  filter(!is.na(NATURALGASD11)) %>%
  group_by(YEAR) %>%
  summarize(mean_consumption = mean(NATURALGASD11), .groups = "drop")

# Bar plot with error bars
ggbarplot(data, x = "YEAR", y = "mean_consumption",
          add = "mean_se",         # Adds mean and standard error bars
          fill = "steelblue",      # Set fill color
          color = "black",         # Outline color
          title = "Average Natural Gas Consumption by Year",
          xlab = "Year", ylab = "Average Consumption") +
  theme_minimal() +
  scale_x_discrete(labels = function(x) substr(x, 3, 4)) # Show last 2 digits of the year


```
```{r}
library(ggpubr)
library(dplyr)

# Prepare data by extracting the year
data <- read.csv("NaturalGasConsumptionMonthlySeries.csv")
data$DATE <- as.Date(data$DATE, format="%m/%d/%Y")
data$YEAR <- format(data$DATE, "%Y")
data <- data %>%
  mutate(NATURALGASD11 = as.numeric(NATURALGASD11)) %>%
  filter(!is.na(NATURALGASD11))

# Box plot by year
ggboxplot(data, x = "YEAR", y = "NATURALGASD11",
          color = "YEAR",      # Color by year for distinction
          palette = "jco",     # Use a color palette
          title = "Distribution of Natural Gas Consumption by Year",
          xlab = "Year", ylab = "Consumption") +
  theme_minimal() +
  scale_x_discrete(labels = function(x) substr(x, 3, 4)) # Display only last 2 digits of the year


```

```{r}
library(ggpubr)
library(dplyr)

# Aggregate data by year for mean consumption
data <- read.csv("NaturalGasConsumptionMonthlySeries.csv")
data$DATE <- as.Date(data$DATE, format="%m/%d/%Y")
data$YEAR <- format(data$DATE, "%Y")
data <- data %>%
  mutate(NATURALGASD11 = as.numeric(NATURALGASD11)) %>%
  filter(!is.na(NATURALGASD11)) %>%
  group_by(YEAR) %>%
  summarize(mean_consumption = mean(NATURALGASD11), .groups = "drop")

# Bar plot with enhanced readability
ggbarplot(data, x = "YEAR", y = "mean_consumption",
          add = "mean_se",
          fill = "steelblue",
          color = "black",
          title = "Average Natural Gas Consumption by Year",
          xlab = "Year", ylab = "Average Consumption") +
  theme_minimal() +
  theme(
    text = element_text(size = 14),             # Increase overall text size
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels
    plot.title = element_text(size = 16, face = "bold") # Bold title
  ) +
  scale_x_discrete(labels = function(x) substr(x, 3, 4)) # Last 2 digits of year


```

```{r}
library(ggpubr)
library(dplyr)

# Prepare data
data <- read.csv("NaturalGasConsumptionMonthlySeries.csv")
data$DATE <- as.Date(data$DATE, format="%m/%d/%Y")
data$YEAR <- format(data$DATE, "%Y")
data <- data %>%
  mutate(NATURALGASD11 = as.numeric(NATURALGASD11)) %>%
  filter(!is.na(NATURALGASD11))

# Box plot with enhanced readability
ggboxplot(data, x = "YEAR", y = "NATURALGASD11",
          color = "YEAR",
          palette = "jco",
          title = "Distribution of Natural Gas Consumption by Year",
          xlab = "Year", ylab = "Consumption") +
  theme_minimal() +
  theme(
    text = element_text(size = 14),             # Increase overall text size
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels
    plot.title = element_text(size = 16, face = "bold") # Bold title
  ) +
  scale_x_discrete(labels = function(x) substr(x, 3, 4)) # Show last 2 digits of year


```

```{r}
library(ggplot2)
library(dplyr)

# Load and prepare data
data <- read.csv("NaturalGasConsumptionMonthlySeries.csv")
data$DATE <- as.Date(data$DATE, format="%m/%d/%Y")
data <- data %>%
  mutate(NATURALGASD11 = as.numeric(NATURALGASD11)) %>%
  filter(!is.na(NATURALGASD11))

# Calculate statistics
mean_consumption <- mean(data$NATURALGASD11)
median_consumption <- median(data$NATURALGASD11)
sd_consumption <- sd(data$NATURALGASD11)
min_consumption <- min(data$NATURALGASD11)
max_consumption <- max(data$NATURALGASD11)

# Calculate trend slope
trend_model <- lm(NATURALGASD11 ~ DATE, data = data)
trend_slope <- coef(trend_model)[2]

# Text annotation for the statistics
stat_text <- paste(
  "Mean:", round(mean_consumption, 2), "\n",
  "Median:", round(median_consumption, 2), "\n",
  "SD:", round(sd_consumption, 2), "\n",
  "Min:", round(min_consumption, 2), "\n",
  "Max:", round(max_consumption, 2), "\n",
  "Trend Slope:", round(trend_slope, 4)
)

# Plot with scatter and smooth line, including the statistics as an annotation
ggplot(data, aes(x = DATE, y = NATURALGASD11)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    title = "Natural Gas Consumption Over Time",
    x = "Date",
    y = "Consumption"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  ) +
  annotate("text", x = min(data$DATE), y = max(data$NATURALGASD11), 
           label = stat_text, hjust = 0, vjust = 1, size = 4)


```


